The Cost of Undisturbed Landscapes

### 2021-01-08 projects/asparagus/doc/asparagus.tex
beginning to rewrite data export to gdx
updates asparagus.tex to version submitted on Dec 8, 2020

\documentclass[review, 3p, times, 12pt]{elsarticle} %final, twocolumn, 5p, 12pt, authoryear, twocolumn, 5p,

@@

% improved float environment
\usepackage{float}
% improved underlines (unterstreichungen)
\usepackage[normalem]{ulem}

% plotting
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.16}

% tables
\usepackage{threeparttable}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{longtable}
\usepackage{tabulary}
\usepackage{threeparttablex}
\usepackage{tabularx}
\usepackage{ltablex}
\usepackage{multirow}

\newcolumntype{M}{>{\raggedright\arraybackslash}X} % Defines new column type with variable length, similar to X in tabularx, but is left aligned rather than justified

%\usepackage{ctable}
\usepackage{makecell}

% SI units
\usepackage{siunitx}
\DeclareSIUnit{\thermal}{th}
\DeclareSIUnit{\electric}{el}
\DeclareSIUnit{\year}{a}
\DeclareSIUnit{\wattpeak}{Wp}
\DeclareSIUnit{\million}{million}

% Euro symbol
\usepackage[gen]{eurosym}

% citations
\usepackage{natbib}

% to-do notes
\setlength{\marginparwidth}{2cm}
\usepackage[draft]{todonotes}

% graphics and figures
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}

\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\COO}{\ensuremath{\mathrm{CO_2}} }

% neater date
\usepackage{etoolbox}
\patchcmd{\MaketitleBox}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}%
\patchcmd{\pprintMaketitle}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}%

\usepackage[colorlinks=true, citecolor=blue, linkcolor=blue, filecolor=blue,urlcolor=blue,pdfauthor=author]{hyperref}

@@

\begin{document}
\begin{frontmatter}
\title{The Cost of Undisturbed Landscapes}
\author[1]{Sebastian Wehrle\corref{cor1}}
\ead{sebastian.wehrle@boku.ac.at}
\author[1]{Johannes Schmidt}
\ead{johannes.schmidt@boku.ac.at}
\author[1]{Christian Mikovits}
\ead{christian.mikovits@boku.ac.at}
\cortext[cor1]{Corresponding author}
\address[1]{Institute for Sustainable Economic Development, University of Natural Resources and Life Sciences,
Feistmantelstrasse 4, 1180 Vienna, Austria}

\begin{abstract}
By 2030 Austria aims to meet \SI{100}{\percent} of its electricity demand from domestic renewable sources, predominantly from wind and solar energy.
While wind power reduces \COO emissions, it is also connected to negative impacts at the local level, such as interference with landscape aesthetics.
Yet, wind power comes at lower system integration cost than solar power, so that it effectively reduces system cost.
We quantify the opportunity cost of replacing wind turbines with solar power, using the power system model \emph{medea}.
Our findings suggest that these cost of undisturbed landscapes are considerable, particularly when PV is not entirely rolled out as utility-scale, open-space installations.
The opportunity cost is likely high enough to allow for significant compensation of the ones affected by local wind turbine externalities.
\end{abstract}

\begin{keyword}
externalities \sep wind power\sep system design\sep policy
\JEL L98 \sep Q42 \sep C61
\end{keyword}
\end{frontmatter}
\newpage

% %%%%% %%%%% %%%%% INTRODUCTION %%%%% %%%%% %%%%% %%%%%
\section{Introduction}\label{sec:introduction}
"Holding the increase in the global average temperature [\ldots] well below \SI{2}{\celsius} above pre-industrial levels and pursuing efforts to limit the temperature increase to \SI{1.5}{\celsius} above pre-industrial levels"~\citep{UNFCCC2015} requires rapid and substantial reductions in greenhouse gas emissions worldwide~\citep{IPCC2018}.
Increasing the share of electricity generated from renewable sources consequently is amongst the United Nations's sustainable development goals (indicator 7.\ 2.\ 1) and a prominent goal of many climate and energy strategies.
The European Union, for example, aims to meet at least \SI{32.5}{\percent} of its final energy consumption from renewable sources~\citep{EU2018}.
Against this background, Austria aims to generate \SI{100}{\percent} of its electricity consumption from domestic renewable sources on annual balance by 2030\footnote{Industry own use and ancillary service provision are exempt.
Together, these accounted for approximately \SI{10}{\percent} of consumption in 2016.}.
According to government estimates, additional electricity generation of \SI{27}{\tera\watt\hour} from renewable
sources is required to meet policy goals.
Out of this total increase in renewable electricity generation, \SI{1}{\tera\watt\hour} is thought to be sourced from biomass, and \SI{5}{\tera\watt\hour} from new hydropower generation, according to the government's climate and
energy strategy~\citep{BMNT2018}.
The comparatively small projected increase in electricity generation from these sources results from the perception
of biomass being ecologically~\citep{Erb2018} and economically unsustainable, while further development of Austria's
hydropower potentials was considered `challenging'~\citep{Wagner2015} even before the projected increase in hydro
generation by \SI{5}{\tera\watt\hour} was factored in.
As Austrians banned the use of nuclear power in a referendum in 1978~\citep{Pelinka1983}, wind and solar
power are the only remaining options for a large-scale expansion of low-carbon power generation in the country.

While wind power mitigates negative external effects at the global (\COO emissions) and the regional (air pollutants
from fossil thermal power generation) level, it also imposes negative external effects at the local level.
Wind turbines are found to negatively impact wildlife~\cite{Loss2013,Voigt2015,Wang2015}, to emit
noise~\cite{Wang2015a}, and to interfere negatively with landscape aesthetics~\citep{Jones2010, Meyerhoff2010}.
The latter is found to be the dominant negative external effect of wind turbines~\cite{Mattmann2016}.

Accordingly, several strands of literature seek to quantify the negative impact of wind turbines, for example
through their impact on property prices within a revealed preferences framework~\citep{Droees2016, Gibbons2015,
Heintzelmann2017, Jensen2018, Kussel2019, Lang2014, Sims2008, Sunak2016, Vyn2014}, or through choice
experiments and surveys~\citep{Drechsler2011, Mattmann2016, Meyerhoff2010} in a stated preferences-context.
Zerrahn~\cite{Zerrahn2017} provides a comprehensive survey of the literature on wind power externalities.

Apart from negative external effects at the local level, wind power is also found to have the potential to reduce
electricity system cost relative to solar PV.
For the case of Germany, Ueckerdt et al.~\cite{Ueckerdt2013} estimate solar PV integration cost being about four times higher than the integration cost of wind power at penetration levels of \SI{25}{\percent}.
Scholz et al. find \cite{Scholz2017} integration cost of \SI{40}[\euro]{\per \mega\watt\hour} for a mix of \SI{80}{\percent} solar PV and \SI{20}{\percent} wind power at \SI{100}{\percent} renewables penetration in Europe.
This reduces to below \SI{25}[\euro]{\per\mega\watt\hour} when the technology mix is changed to \SI{20}{\percent} solar PV and \SI{80}{\percent} wind power.
% Similarly, Gils et al. \cite{Gils2017}

These findings illustrate that abandoning wind power in favor of undisturbed landscapes potentially comes at significant opportunity cost.
We complement the literature on negative wind turbines impacts and set out to quantify this cost of undisturbed landscapes for the case of Austria.
First, Austrian policies targeting at a fixed share of renewable electricity generation are an ideal framework for
studying the effects of substituting renewable energy generation technologies (RET).
Moreover, Austria has large hydro storage capacities (including seasonal hydro storages) in place, which should limit PV integration cost.
Thus, our findings should be qualitatively generalizable to power systems with high renewable shares and lower storage capacities in similar climates.
Methodologically, we rely on the power system model \emph{medea}, which is described in section~\ref{sec:data-and-methods}.

Our analysis contributes to a more comprehensive understanding of the trade-offs faced in the design of highly renewable electricity systems, thereby helping to inform policies about the socially optimal expansion of RET.
Any such policy necessarily needs to account for the full social cost (including all externalities) of RET.
Moreover, we contribute to the analysis of near-optimal power system configurations~\citep[see e.g.][]{Neumann2019,
Schlachtberger2017}.
Corresponding results are presented in section~\ref{sec:results-discussion}.

Our results also provide insight into how potential compensation for negative external effects of wind turbines at the local level could be financed by system-wide benefits of wind power.
Moreover, based on our analysis we can approximate the valuation of undisturbed landscapes that is implicit in Austria's climate and energy strategy.
These policy implications and conclusions are summarized in section~\ref{sec:conclusions-policy-implication}.

@@

\section{Data and Methods} \label{sec:data-and-methods}
To investigate the cost of undisturbed landscapes (i.e. the opportunity cost of wind power) we use the power system model \emph{medea}, as summarized in section~\ref{subsec:medea}.
We instantiate the model with scenario assumptions and data as observed in the year 2016, which was a low-wind year in Austria (see section~\ref{subsec:data}).

Based on this, we approximate the opportunity cost of wind power through the following procedure:
\begin{enumerate}
\item We derive the unrestricted system cost-minimizing deployment of wind and solar power, given the scenario
assumptions
\item We restrict deployment of wind power by a small margin (so that the next best RET substitutes for wind
power) and observe net system cost $c_{net}$ for Austria (calculated as total system cost including air
pollution cost net of trade).\label{enumerate:restrict}
\item We repeat step~\ref{enumerate:restrict} till no wind power can be deployed.
\item Finally, we approximate the opportunity cost of wind power (at given wind power capacity $w$) $OC_w$ by
the change in $c_{net}$ in response to a change in wind power capacity $w$ deployed, i.e.
\[OC_w = \frac{\Delta c_{net}}{\Delta w}\]
where $\Delta$ is the difference operator and opportunity cost $OC_w$ are expressed in \EUR per
\SI{}{\mega\watt} wind power foregone.
\end{enumerate}

% ***** medea *****
\subsection{Power system model \emph{medea}}\label{subsec:medea}
We make use of the power system model \emph{medea} to simulate (dis)investment in and hourly operation of the
prospective Austrian (and German) power system.
The model is cast as a linear optimization seeking to minimize total system cost, which consist of fuel and emission
cost, quasi-fixed and variable operation and maintenance (O\&M) cost, the costs of investment in energy generation,
storage, and transmission assets, and (potential) cost of non-served load.
From an economic perspective, the model reflects a perfectly competitive energy-only market with fully
price-inelastic demand and perfect foresight of all actors.

The system is required to meet exogenous and inelastic demand for electricity and heat in any hour of the modeled
year.
Energy supply, in turn, is constrained by available installed capacities of energy conversion, storage, and
transmission units.
Co-generation units convert fuel to heat and power subject to a feasible operating region defined by the unit's
electrical efficiency, the electricity loss per unit of heat production, and the back pressure coefficient.
Electricity generation from intermittent sources (wind, run-of-river hydro, solar) is subject to exogenous hourly
generation profiles, which are scaled according to total installed capacities.
Electricity from these sources can be curtailed at no additional cost (free disposal).

Electricity can be stored in reservoir and pumped hydro storages or batteries.
The capacity of hydro storages can not be expanded, as we assume existing potentials to be exhausted.
Battery capacity, on the other hand, can be added endogenously.
Generation from storages is constrained by installed capacity and energy contained in storage.
Hydro reservoirs are filled (exogenous) inflows.
Pumped hydro storages and batteries can actively store electricity for later use.
To better capture operational differences of hydro storage units, we model daily, weekly and seasonal reservoir and pumped storage plant separately.
To ensure a stable and secure operation of the electricity system, ancillary services (e.g. frequency control,
voltage support) are required.
We model ancillary service needs as a minimum requirement on spinning reserves operating at any point in time.
Thus, we assume that ancillary services can be provided by thermal power plant, run-of-river hydro plant or any
type of storage technology.
We do not model transmission and distribution grids within Austria or Germany.
We do, however, allow for cross-border electricity trade between both countries.
We also keep track of the cost of air pollution arising from burning fossil fuels.
As these cost are external, they do not enter cost minimization.
For a detailed, mathematical description of the model please refer to \ref{sec:medea-desc}.
Data processing is implemented in python, while the optimization model is based on GAMS.
Running the model in hourly resolution for one year takes 10--15 minutes, depending on the model parameters, on an Intel i7-8700 machine with 16 GB RAM, using CPLEX 12.9 as a solver.

% ***** Data *****
\subsection{Data}\label{subsec:data}
We set up our model with the goal to resemble Austria's prospective electricity and district heating systems in the year 2030.
We also include Austria's largest electricity trading partner, Germany, to account for potential effects from electricity trade.
Again, our scenario reflects Germany's currently announced electricity sector policies, so that we set generation capacities to levels consistent with policies effective by 2030.
These assumptions are laid out in sections~\ref{subsubsec:assumptions-austria} and \ref{subsubsec:assumptions-germany} and summarized in table~\ref{tab:instcaps}.

\subsubsection{Scenario Assumptions for Austria}\label{subsubsec:assumptions-austria}
The Austrian government has set itself the goal of generating \SI{100}{\percent} of electricity consumption from domestic renewable sources on annual balance by 2030~\citep{Regierungsprogramm2020}.
However, industry own consumption and system services, which currently account for about \SI{10}{\percent} of annual electricity consumption, are exempt.

The government plans to achieve this goal by generating an additional \SI{27}{\tera\watt\hour} of electricity annually from renewable sources\footnote{Please note that this policy goal does not imply that the generated
electricity must actually be consumed. Hence, we count curtailed electricity as contributing to the policy goal.}.
New hydro power plants are thought to contribute \SI{5}{\tera\watt\hour} annually, while additional electricity generation from biomass is envisaged to account for \SI{1}{\tera\watt\hour} annually.
In our scenarios, we add generation capacities sufficient for reaching these targets.

Further, the government projects the remainder to come from solar PV (\SI{11}{\tera\watt\hour} annually) and onshore wind turbines (\SI{10}{\tera\watt\hour} annually)~\cite{Regierungsprogramm2020}.
As we are interested in determining the opportunity cost of wind power versus its best alternative, we allow for endogenous investment in wind and solar power, without enforcing announced targets.
Other low-carbon energy technologies are not feasible at large scale in the Austrian context, as we have laid out in the introduction.
Initial generation capacities for Austria are summarized in table~\ref{tab:instcaps}.

\subsubsection{Scenario Assumptions for Germany}\label{subsubsec:assumptions-germany}
Germany has announced specific capacity targets for several power generation technologies.
Following these announcements, we anticipate an end to nuclear power generation, a (partial) coal exit according to
recommendations by the `coal commission'~\cite{WSB2019}, and a further expansion of renewable electricity generation
in line with the German Renewable Energy Sources Act~\cite{noauthor_renewable_2017}.
We do, however, expect the \SI{52}{\giga\watt} solar PV cap to be removed.
The corresponding capacity assumptions are displayed in table~\ref{tab:instcaps}.

\begin{table}[H]
\caption{Initial Generation Capacities} \label{tab:instcaps}
\centering
\begin{tabulary}{0.55\textwidth}{L C C}
\toprule
Technology & Austria & Germany\\
& \si{\giga\watt} & \si{\giga\watt}\\
\midrule
Wind Onshore & $2.6$ & $90.8$ \\
Wind Offshore  & $0.0$ & $15.0$ \\
Solar PV & $1.1$ & $73.0$ \\
Run-of-river Hydro  & $6.9$ & $4.5$ \\
Hydro storage & $7.7$ & $6.5$ \\
Biomass & $0.7$ & $8.4$ \\
Lignite & $0.0$ & $11.4$ \\
Coal & $0.0$ & $14.0$ \\
Natural Gas & $3.9$ & $24.2$ \\
Mineral Oil & $0.2$ & $3.5$ \\
Heat Pump & $1.0$ & $1.0$ \\
Gas Boiler & $3.9$ & $25.5$\\ \bottomrule
\end{tabulary}
\end{table}

\subsubsection{Energy supply}
We represent 20 dispatchable energy conversion and storage technologies that are expected to be operated in 2030.
In addition to given, initially installed capacities (see table~\ref{tab:instcaps}), the model can endogenously add further generation capacities that are compatible with stated policy objectives.
We calculated annualized investment cost of each technology on the basis of an assumed weighted average cost of capital of \SI{5}{\percent} over the plant's lifetime.
All technologies can also be decommissioned, so that we adopt a long-run perspective on the power system.
Parameters of admissible technologies are summarized in Table~\ref{itm:techcost}.

>>>>

\documentclass[review, 3p, times, 12pt, authoryear]{elsarticle}

@@

\usepackage{float}
\usepackage[normalem]{ulem}

\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.16}

\usepackage{threeparttable}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{longtable}
\usepackage{tabulary}
\usepackage{threeparttablex}
\usepackage{tabularx}
\usepackage{ltablex}
\usepackage{multirow}

\newcolumntype{M}{>{\raggedright\arraybackslash}X}
\usepackage{makecell}

\usepackage{siunitx}
\DeclareSIUnit{\wattth}{Wth}
\DeclareSIUnit{\wattel}{We}
\DeclareSIUnit{\year}{a}
\DeclareSIUnit{\wattpeak}{Wp}
\DeclareSIUnit{\million}{million}
\DeclareSIUnit{\coo}{\ce{CO2}e}

\usepackage[gen]{eurosym}
\usepackage[version=4]{mhchem}
\defcitealias{BMNT2018}{BMNT and BMVIT, 2018}
\defcitealias{Regierungsprogramm2020}{BKA, 2020}
\defcitealias{WSB2019}{WSB, 2019}
\defcitealias{ENTSOE2020a}{ENTSO-E, 2020a}
\defcitealias{ENTSOE2020b}{ENTSO-E, 2020b}
\defcitealias{AGEB2018}{AGEB, 2018}
\defcitealias{CCCS2017}{C3S, 2017}
\defcitealias{IMF2020}{IMF (2020)}
\defcitealias{opsd2019}{OPSD (2019)}

\usepackage{graphicx}
\graphicspath{{./figures/}}
\usepackage{caption}
\usepackage{subcaption}

\DeclareMathOperator*{\argmax}{arg\,max}

\usepackage{etoolbox}
\patchcmd{\MaketitleBox}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}
\patchcmd{\pprintMaketitle}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}

\usepackage[hidelinks, colorlinks=false, citecolor=blue, linkcolor=blue, filecolor=blue,urlcolor=blue,pdfauthor=Sebastian Wehrle]{hyperref}

\newcommand*{\figref}[2][]{%
    \hyperref[{fig:#2}]{%
        Figure~\ref*{fig:#2}%
        \ifx\\#1\\%
        \else
        \,#1%
        \fi
    }%
}

@@

\begin{document}
    \begin{frontmatter}
        \title{The Cost of Undisturbed Landscapes}
        \author[1]{Sebastian Wehrle\corref{cor1}}
        \ead{sebastian.wehrle@boku.ac.at}
        \author[1]{Johannes Schmidt}
        \ead{johannes.schmidt@boku.ac.at}
        \cortext[cor1]{Corresponding author}
        \address[1]{Institute for Sustainable Economic Development, University of Natural Resources and Life Sciences,
        Feistmantelstrasse 4, 1180 Vienna, Austria}

        \begin{abstract}
            By 2030, Austria aims to meet \SI{100}{\percent} of its electricity demand from domestic renewable sources, such as wind energy.
            Apart from the benefit of reducing \ce{CO2} emissions and, potentially, system cost, wind power is also associated with negative impacts at the local level, particularly its interference with landscape aesthetics.
            Some of these negative impacts might be avoided by using alternative renewable energy technologies.
            To better understand the trade-offs involved, we quantify the opportunity cost of wind power versus its best feasible alternative, which is solar photovoltaics in Austria. Methodologically, our analysis relies on the power system model \emph{medea}, a technology-rich, partial-equilibrium model of the Austro-German electricity and district heating markets.
            Our findings suggest that the cost of undisturbed landscapes is considerable, particularly when solar PV is mainly realized on rooftops.
            Under a wide range of assumptions, the opportunity cost of wind power is high enough to allow for significant compensation of the ones affected by local, negative wind turbine externalities.
        \end{abstract}

        \begin{keyword}
            externalities \sep wind power\sep system design\sep policy
            \JEL L98 \sep Q42 \sep C61
        \end{keyword}
    \end{frontmatter}
    \newpage
% %%%%% %%%%% %%%%% INTRODUCTION %%%%% %%%%% %%%%% %%%%%

    \section{Introduction}\label{sec:introduction}
    "Holding the increase in the global average temperature [\ldots] well below \SI{2}{\celsius} above pre-industrial levels and pursuing efforts to limit the temperature increase to \SI{1.5}{\celsius} above pre-industrial levels"~\citep{UNFCCC2015} requires rapid and substantial reductions in greenhouse gas emissions worldwide~\citep{IPCC2018}.
    Increasing the share of electricity generated from renewable sources consequently is amongst the United Nations' sustainable development goals (indicator 7.2.1) and a prominent goal of many climate and energy strategies.
    The European Union, for example, aims to meet at least \SI{32}{\percent} of its final energy consumption from renewable sources by 2030~\citep{EU2018}.\footnote{In its "European Green Deal" the EU Commission proposes to increase ambitions and reduce \ce{CO2} emissions by at least \SI{55}{\percent} by 2030. This compares to a \SI{40}{\percent} reduction target previously and is likely to have knock-on effects on the expansion of renewable energy generation.}
    Against this background, Austria aims to generate \SI{100}{\percent} of its electricity consumption from domestic renewable sources on annual balance by 2030\footnote{Industry's own use and ancillary service provision are exempt. Together, these accounted for approximately \SI{10}{\percent} of consumption in 2016.}.
    The required expansion of renewable energy generation shall be incentivized by technology-specific subsidies rather than \ce{CO2} pricing.

    According to government estimates, an additional \SI{27}{\tera\watt\hour} of electricity needs to be generated from renewable sources annually to meet this policy goal.
    Out of the targeted total increase in renewable electricity generation, \SI{1}{\tera\watt\hour} is thought to be sourced from biomass, and \SI{5}{\tera\watt\hour} from new hydropower generation, according to the government's climate and energy strategy~\citepalias{BMNT2018}.
    The comparatively small projected increase in electricity generation from these sources results from the perception of biomass being ecologically\footnote{More specifically, increasing biomass harvests would decrease carbon stocks strongly, both through deforestation and management. Thus, a net positive greenhouse gas balance of bioenergy would only be possible after a rather long payback time. However, additions to the stock of atmospheric greenhouse gases need to be reduced rapidly.}
    ~\citep{Erb2018} and economically unsustainable\footnote{The temporary expiration of Austria's subsidy scheme for biomass plants in 2019 resulted in a widespread shutdown of biomass-fired generators.}
    , while further development of Austria's hydropower potential was considered `challenging'~\citep{Wagner2015} even before the projected \SI{5}{\tera\watt\hour} increase in hydro generation was factored in.
    As Austrians banned the use of nuclear power in a referendum in 1978~\citep{Pelinka1983}, wind and solar power are the only remaining options for a large-scale expansion of low-carbon power generation in the country.

    While wind power mitigates negative external effects at the global (\ce{CO2} emissions) and the regional (air pollutants from fossil thermal power generation) level, it also imposes negative external effects at the local level.
    Wind turbines are found to impact wildlife negatively~\citep{Loss2013,Voigt2015,Wang2015} and to emit noise~\citep{Wang2015a}.
    They are also associated with land-use and land-cover change~\citep{Turkovska2020} and with negative interference with landscape aesthetics in general~\citep{Jones2010, Meyerhoff2010}.
    The latter is found to be the dominant, negative external effect of wind turbines~\citep{Mattmann2016}.
    Accordingly, several strands of literature seek to quantify the negative impact of wind turbines, for example through their effect on property prices within a revealed preference framework~\citep{Droees2016, Gibbons2015,
    Heintzelmann2017, Jensen2018, Kussel2019, Lang2014, Sims2008, Sunak2016, Vyn2014}, or through choice experiments and surveys~\citep{Drechsler2011, Mattmann2016, Meyerhoff2010} in a stated preferences-context.
    \cite{Zerrahn2017} provides a comprehensive survey of the literature on wind power externalities.

    Apart from negative external effects at the local level, wind power is also found to have the potential to reduce electricity system costs, in particular, relative to solar PV.
    For the case of Germany, \cite{Ueckerdt2013} estimate solar PV integration cost to be about four times higher than the integration cost of wind power at penetration levels of \SI{25}{\percent}.
    \cite{Scholz2017} find integration cost of \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{40}[\euro]{\per\mega\watt\hour} for a mix of \SI{80}{\percent} solar PV and \SI{20}{\percent} wind power at \SI{100}{\percent} renewables penetration in Europe.
    This reduces to below \SI[per-mode=symbol,sticky-per, bracket-unit-denominator=false]{25}[\euro]{\per\mega\watt\hour} when the technology mix consists of \SI{20}{\percent} solar PV and \SI{80}{\percent} wind power.
    These findings illustrate that abandoning wind power for PV in favor of undisturbed landscapes potentially comes at significant opportunity cost.
    We complement the literature on negative wind turbine impacts and set out to quantify this cost of undisturbed landscapes for the case of Austria.
    First, Austrian policies targeting a fixed share of renewable electricity generation are an ideal framework for studying the effects of substituting renewable energy generation technologies (RET).
    Moreover, Austria has large hydro storage capacities (including seasonal hydro storages) in place, which should limit PV integration cost.
    Thus, our findings should be qualitatively generalizable to power systems with high renewable shares but lower storage capacities with climates comparable to Austria.
    Methodologically, we rely on the power system model \emph{medea}, which is described in \autoref{sec:data-and-methods}.

    Our analysis contributes to a more comprehensive understanding of the trade-offs faced in the design of highly renewable electricity systems, thereby helping to inform policies about the socially optimal expansion of RET.
    Any such policy necessarily needs to account for the full social cost, including all externalities, of RET.
    Further, we contribute to the analysis of near-optimal power system configurations~\citep[see e.g.][]{Neumann2019, Schlachtberger2017}.
    Corresponding results are presented in \autoref{sec:results-discussion}.
    Finally, our investigation into the cost of undisturbed landscapes generates insight into the possibilities and limitations of funding potential compensating measures for the benefit of the ones negatively affected by wind turbines.
    We illustrate these potentials by an approximation of the valuation of undisturbed landscapes that is implicit in Austria's climate and energy strategy.
    \hyperref[sec:conclusions-policy-implication]{Section~\ref*{sec:conclusions-policy-implication}} summarizes policy implications and concludes.

@@

    \section{Data and Methods} \label{sec:data-and-methods}
    To investigate the cost of undisturbed landscapes (i.e. the opportunity cost of wind power) we use the power system model \emph{medea}, as summarized in \autoref{subsec:medea}.
    We instantiate the model with scenario assumptions and data as observed in 2016, which was a low-wind year in Austria, while solar and hydro resources were close to long-term averages (see \autoref{subsec:data}).
    Based on this, we approximate the opportunity cost of wind power through the following procedure:
    \begin{enumerate}
        \item We derive the unrestricted system cost-minimizing deployment of wind and solar power, given the scenario assumptions
        \item We restrict deployment of wind power by a small margin (so that PV, the next best RET, substitutes for wind power) and observe net system cost $c_{net}$ for Austria, calculated as total system cost including air pollution cost net of the balance of trade.\label{enumerate:restrict}
        \item We repeat step~\ref{enumerate:restrict} till no wind power can be deployed.
        \item Finally, we approximate the opportunity cost of wind power (at given wind power capacity $w$) $OC_w$ by the change in $c_{net}$ in response to a change in wind power capacity $w$ deployed, i.e.\ \[OC_w = \frac{\Delta c_{net}}{\Delta w}\]
        where $\Delta$ is the difference operator and opportunity cost $OC_w$ are expressed in \EUR per \SI{}{\mega\watt} wind power foregone.
    \end{enumerate}

% ***** medea *****

    \subsection{Power system model \emph{medea}} \label{subsec:medea}
    We make use of the power system model \emph{medea} to simulate (dis)investment in, and hourly operation of, the prospective Austrian and German power systems in the year 2030.
    The model is cast as a linear optimization seeking to minimize total system cost, which consists of fuel and emission costs, quasi-fixed and variable operation and maintenance (O\&M) costs, the costs of investment in energy generation, storage, and transmission assets, and the (potential) cost of non-served load.
    From an economic perspective, the model reflects a perfectly competitive energy-only market with fully price-inelastic demand and perfect foresight of all actors.

    The modeled system is required to meet exogenous and inelastic demand for electricity and heat at any hour of the year.
    Energy supply, in turn, is constrained by available installed capacities of energy conversion, storage, and transmission units.
    Co-generation units convert fuel to heat and power subject to a feasible operating region defined by the unit's electrical efficiency, the electricity loss per unit of heat production, and the backpressure coefficient.
    Electricity generation from intermittent sources (wind, run-of-river hydro, solar) is subject to exogenous hourly generation profiles, which are scaled according to total installed capacities.
    Electricity from these sources can be curtailed at no additional cost (free disposal).

    Electricity can be stored in reservoir and pumped hydro storages, batteries, or hydrogen storage (more precisely: water electrolysis, hydrogen storage, and reconversion to electricity in fuel cells).
    The capacity of hydro storages can not be expanded, as we assume existing potentials to be exhausted.
    Battery and hydrogen storage capacities, on the other hand, can be added endogenously.
    Generation from storage is constrained by installed capacity and stored energy.
    Inflows of water into reservoirs add to stored energy.
    Pumped hydro storages, batteries, and hydrogen storage can actively store electricity for later use.
    To better capture operational differences of hydro storage units, we model daily, weekly and seasonal reservoir and pumped storage plants separately.
    To ensure the stable and secure operation of the electricity system, ancillary services (e. g. frequency control, voltage support) are required.
    We model ancillary service needs as a minimum requirement on spinning reserves operating at any point in time.
    Thus, we assume that ancillary services can be provided by thermal power plants, run-of-river hydro plants, or any (active) storage technology.
    We don't model transmission and distribution grids within Austria or Germany.
    We do, however, allow for cross-border electricity trade between both countries.
    We also keep track of the cost of (non-\ce{CO2}) air pollution arising from burning fossil fuels.
    As these costs are external, they do not enter cost minimization.
    For a detailed mathematical description of the model, please refer to \ref{sec:medea-desc}.
    Data processing is implemented in python, while the optimization model is based on GAMS. The model code is published at \url{https://github.com/inwe-boku/medea} under an open MIT license.
    Running the model in hourly resolution for one year takes 10-15 minutes, depending on the model parameters, on an Intel i7-8700 machine with 16 GB RAM, using CPLEX 12.10 as a solver.

% ***** Data *****

    \subsection{Data} \label{subsec:data}
    We set up our model with the goal to resemble Austria's prospective electricity and district heating systems in the year 2030.
    We also include Austria's largest electricity trading partner, Germany, to account for potential effects from electricity trade.
    For Germany, our scenario reflects current and announced future electricity sector policies up to 2030, so that we set generation capacities to levels consistent with these policies.
    These assumptions are laid out in subsections \ref{subsubsec:assumptions-austria} and \ref{subsubsec:assumptions-germany}, and summarized in \autoref{tab:instcaps}.

    \subsubsection{Scenario Assumptions for Austria} \label{subsubsec:assumptions-austria}
    The Austrian government has set itself the goal of generating \SI{100}{\percent} of electricity consumption from domestic renewable sources on annual balance by 2030~\citepalias{Regierungsprogramm2020}.
    However, industry own consumption and system services, which currently account for about \SI{10}{\percent} of annual electricity consumption, are exempt.
    The government plans to achieve this goal by generating an additional \SI{27}{\tera\watt\hour} of electricity annually from renewable sources\footnote{Please note that this policy goal does not imply that the generated electricity must actually be consumed. Hence, we count curtailed electricity as contributing to the policy goal.}
    , for a total of \SI{78.1}{\tera\watt\hour} of renewable electricity generation in 2030.
    New hydropower plants are thought to contribute \SI{5}{\tera\watt\hour} annually, while additional electricity generation from biomass is envisaged to account for \SI{1}{\tera\watt\hour} annually.
    In our scenarios, we add generation capacities sufficient to reach these targets.
    Further, the government projects the remainder to come from solar PV and onshore wind turbines with an annual contribution of \SI{11}{\tera\watt\hour} and \SI{10}{\tera\watt\hour} \citepalias{Regierungsprogramm2020}, respectively.

    As we are interested in determining the opportunity cost of wind power versus its best alternative, we allow for endogenous investment in wind and solar power without enforcing announced technology-specific targets.
    Other low-carbon energy technologies are not feasible at a large scale in the Austrian context, as we have laid out in the introduction.
    Initial generation capacities for Austria are summarized in \autoref{tab:instcaps}.
